\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-09-18 paper/draft.tex
Update intro

\section{Introduction}

The development of automatic disease prediction systems 
is crucial for enhancing healthcare and supporting medical 
professionals in timely and accurate diagnosis. 
Symbolic artificial intelligence (AI), particularly ASP, 
has been explored in this context due to its ability 
to support structured and explainable reasoning. 
However, the construction of expressive knowledge bases 
remains a significant barrier to practical adoption.

This paper introduces \sys, a framework 
that combines LLMs with the 
formal reasoning capabilities of ASP.
\sys extracts and encodes medical knowledge from text into ASP code,
enabling interpretable and accurate diagnosis 
of diseases based on patient symptoms.

\section{Background}

>>>>

\section{Introduction}
Getting an accurate diagnosis quickly can make all the difference in patient outcomes, 
but building automated systems that practitioners can actually trust is still a major challenge. 
Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning \cite{Brewka2011,Gelfond_Kahl_2014}. 
However, its adoption has been limited because constructing and maintaining large, 
domain-specific knowledge bases is very labor-intensive.
On the other hand, Large language models (LLMs) achieve strong results on tasks
such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use \cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.

In clinical decision support, the real goal is to get the best of both worlds: 
systems that reason accurately while still being easy for practitioners to trust and use. 
\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
Instead of requiring experts to hand-craft large knowledge bases, 
\sys automatically converts medical literature into ASP rules. 
These rules are then enriched with patient symptoms and clinical indicators, 
and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.

Concretely, this work:
\begin{itemize}
    \item Introduces a framework that turns medical text into an executable ASP knowledge base.
    \item Uses prompts to generate modular rules (including alternatives and symptom links) that the solver applies for reasoning.
    \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
    \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
\end{itemize}

On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
while giving transparent explanations for each diagnosis.

The paper begins with background on LLMs and ASP (\cref{sec:background}), 
then details the \sys methodology (\cref{sec:methodology}), 
evaluates the approach on patient data (\cref{sec:results}), 
and concludes with a discussion of limitations and future directions (\cref{sec:conclusion}).

\section{Background}\label{sec:background}
